### 2017-07-04 motivation.tex
edits

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in this language.  TPC-C is a
well-known Online Transaction Processing (OLTP) benchmark that models
an order-processing system for a wholesale parts supply business. The
business logic is captured in 5 database transactions that operate on
9 tables; \C{new\_order} is one such transaction that uses
\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
\C{Order\_line} tables. The transaction acts on the behalf of a
customer, whose id is \C{c\_id}, to place a new order for a given
set of items (\C{item\_reqs}), to be served by a warehouse under the
district identified by \C{d\_id}.  Fig.~\ref{fig:scheme} illustrates
the relationship among these different tables.

@@

The \C{new\_order} transaction inserts a new \C{Order} record, whose
id is the sequence number of the next order under the given district
(\C{d\_id}). The sequence number is stored in the corresponding
\C{District} record, and updated each time a new order is added to the
system. Since each order may request multiple items (\C{item\_reqs}),
an \C{Order\_line} record is created for each requested item to relate
the order with the item. Each item has a corresponding record in the
\C{Stock} table, which keeps track of the quantity of the item left in
stock (\C{s\_qty}). The quantity is updated by the transaction to
reflect the processing of new order (if the stock quantity falls below
10, it is automatically replenished by 91).

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verfiying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation does not violate
  atomicity as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs (e.g., garbage
collectors~\cite{JLP+14,GHE15,HPQ+15}), weak isolation introduces new
challenges arising from the use of transactions, and new opportunities
arising from the fact that the store abstraction used by transactions
is a relational database, not low-level memory.

@@

This simple characterization of RR isolation allows us to verify the
consistency conditions associated with the \C{new\_order}
transaction. First, since the database doesn't change ($\Delta' =
\Delta$) (where $\Delta'$ represents the state after the transaction
commits), during the execution of the transaction's body, we can
reason about \C{new\_order} as though it executed in complete isolation
until its commit point, leading to a verification process similar to
what would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution and becomes ready to commit, a
transition that transfers the transaction's local writes ($\delta$) to
the unchanged database state ($\Delta$).  However, we are required to
consider the interference from concurrent transitions at this point,
which might change the database state from $\Delta$ to $\Delta_c$. If
this interference includes the effects of a concurrent \C{new\_order}
transaction (with the same \C{d\_id}), then verification fails as
described previously (Fig.~\ref{fig:new_order_execs}). Fortunately,
sequential reasoning shows that this is impossible - RR prevents a
concurrent \C{new\_order} transaction that modifies the same
\C{District} record as the current transaction (concretely, since the
record is already present in the current transaction's local log, any
transition from $\Delta$ to $\Delta_c$ cannot change this record).
Applying such axiomatic reasoning on \C{new\_order} allows us to prove
that the TPC-C invariant holds when the transaction is executed under
PostgreSQL's RR isolation.  Our proof framework generalizes this style
of reasoning to various isolation levels on databases.

@@

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically. We find it useful to reason about
statements that mutate the database state, not in terms of a pre- and
post-condition pair, but in terms of a state transformer that relates
the pre- and post-states of a statement. This state transformer
semantics can be defined algorithmically, just like predicate
transformer semantics (e.g., strongest post-condition).  Here, a state
transformer interprets a SQL statement in the set domain, taking
advantage of the fact that a a database is essentially a set of
tuples, and a SQL statement is a transformer over these sets.  The
benefit of this approach is that low-level loops can now be
substituted with higher-order combinators that automatically lift the
state transformer of its higher-order argument (i.e., the loop body)
to the state transformer of the combined expression (i.e., the loop).
Thus the semantics of a \C{foreach} loop, for instance, can be
captured as a state transformer, where the state is a set, and the
transformation defines a bind operation. We illustrate this intuition
on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as following:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariants, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The following two sections,
however, first focus on laying the theoretical foundations for
reasoning about weak isolation.

>>>>

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in this language.  TPC-C is a
well-known Online Transaction Processing (OLTP) benchmark that models
an order-processing system for a wholesale parts supply business. The
business logic is captured in 5 database transactions that operate on
9 tables; \C{new\_order} is one such transaction that uses
\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
\C{Order\_line} tables. The transaction acts on the behalf of a
customer, whose id is \C{c\_id}, to place a new order for a given
set of items (\C{item\_reqs}), to be served by a warehouse under the
district identified by \C{d\_id}.  Fig.~\ref{fig:schema} illustrates
the relationship among these different tables.

@@

The \C{new\_order} transaction inserts a new \C{Order} record, whose
id is the sequence number of the next order under the given district
(\C{d\_id}). The sequence number is stored in the corresponding
\C{District} record, and updated each time a new order is added to the
system. Since each order may request multiple items (\C{item\_reqs}),
an \C{Order\_line} record is created for each requested item to relate
the order with the item. Each item has a corresponding record in the
\C{Stock} table, which keeps track of the quantity of the item left in
stock (\C{s\_qty}). The quantity is updated by the transaction to
reflect the processing of new orders (if the stock quantity falls below
10, it is automatically replenished by 91).

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation does not violate
  atomicity as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs (e.g., garbage
collectors~\cite{JLP+14,GHE15,HPQ+15}), weak isolation introduces new
challenges arising from the use of transactions, and new opportunities
arising from the fact that the store abstraction used by transactions
is a relational database, not low-level memory.

@@

This simple characterization of RR isolation allows us to verify the
consistency conditions associated with the \C{new\_order}
transaction. First, since the database doesn't change ($\Delta' =
\Delta$) during the execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution and becomes ready to commit, a
transition transfers the transaction's local writes ($\delta$) to the
unchanged database state ($\Delta$).  However, we are required to
consider the interference from concurrent transitions at this point,
which might change the database state from $\Delta$ to $\Delta_c$. If
this interference includes the effects of a concurrent \C{new\_order}
transaction (with the same \C{d\_id}), then verification fails as
described previously.  Fortunately, sequential reasoning shows that
this is impossible - RR prevents a concurrent \C{new\_order}
transaction that modifies the same \C{District} record as the current
transaction (concretely, since the record is already present in the
current transaction's local log, any transition from $\Delta$ to
$\Delta_c$ cannot change this record).  Applying such axiomatic
reasoning on \C{new\_order} allows us to prove that the TPC-C
invariant holds when the transaction is executed under PostgreSQL's RR
isolation.  Our proof framework generalizes this style of reasoning to
various isolation levels on databases.

@@

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically. We find it useful to reason about
statements that mutate the database state, not in terms of a pre- and
post-condition pair, but in terms of a state transformer that relates
the pre- and post-states of a statement. This state transformer
semantics can be defined algorithmically, just like predicate
transformer semantics (e.g., strongest post-condition).  Here, a state
transformer interprets a SQL statement in the set domain, taking
advantage of the fact that the database is essentially a set of
tuples, and a SQL statement is a transformer over these sets.  The
benefit of this approach is that low-level loops can now be
substituted with higher-order combinators that automatically lift the
state transformer of its higher-order argument (i.e., the loop body)
to the state transformer of the combined expression (i.e., the loop).
Thus the semantics of a \C{foreach} loop, for instance, can be
captured as a state transformer, where the state is a set, and the
transformation defines a bind operation. We illustrate this intuition
on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as follows:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariants, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The following two sections,
however, first focus on laying the theoretical foundations for
reasoning about weak isolation.
